% People: Ian, Andy, Jack, India, Mum, Dad, Lee, Jo, Claire, Vickie, Chris, Akash, Nick, Alex, Jake, Billy, George, Ricky, Muntazim, Connor W, Raf, Dan, Joe, Gareth, Ronaldas, Hooshyar, Christina, Toyah, Poco, Xan 

% Examiners
Firstly, I owe special thanks to Graham and Kazuya for examining my thesis and conducting my viva voce.
%
% Ian, Andy, Gareth, Xan
Thank you to my academia supervisors: Andy Lundgren, you gave me the first opportunity to perform some `proper' research; Ian Harry, you supported me throughout the PhD as my main supervisor, ensuring I was suitably challenged and consequently over-prepared all the way to the end; Gareth and Xan you have both helped me immeasurably with my projects and saved me from countless hours of debugging.

% Vickie
I required a great deal of personal support during the PhD, and for that the bulk of thanks goes to my amazing partner, Vickie. Without you by my side, I'm sure I wouldn't have made it to the end. There were many moments I was ready to quit and get a job that paid me a living wage. Thankfully, you were there to help me out and maintain my lavish lifestyle, I love you forever. 

% Friends
% Billy, George, Ricky, Muntazim, Harrison, Luke
% Chris, Akash, Nick, Alex, Jake
% Connor, Rafaela, Joe, Dan
% Molly, Susanna, Rahma
% Jinglan
% Ronaldas
To my friends in Ipswich: Billy, George, Ricky, Muntazim, I am thankful for the friendship we've shared throughout primary school, high school and sixth form. To the friends I made during my undergraduate degree: Chris, Akash, Nick and Alex, we've had great adventures and living together made University feel like a breeze, I truly wish I could experience it all again. I want to say special thanks to Chris for being an amazing flatmate during the initial years of the PhD, I miss the days when we would spend hours chatting in the kitchen.

Thanks to the many PhD students I have shared a department with, you made this journey far less daunting that it could have been, foremost are my cohort: Connor, Rafaela, Dan, and Joe. Connor, we spent hours ranting about the PhD, you were the best support a man could've asked for and knowing we weren't alone in how we felt meant we both made it to the end, and it wasn't all for nothing. Love you buddy.

% Family
% India, Jack
% Mum, Lee
% Dad, Jo
% Claire
The root cause of my success is my family. These people have moulded me into the person I am, and I love you all more than anything. Jack and India, thank you for the childhood we shared. Mum and Lee, thank you for your constant support and safety net. Dad, I am thankful for who you are today and I know you're always trying your best, Jo, thank you for forcing him to try his best! Claire, thank you for always doing the best for me, no matter the circumstances.
